Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-07 text/manuscript-full-KS.qmd
revisions rq2 Chicago

@fig-chicago-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Chicago metropolitan area, notably revealing that the City of Chicago stands out distinctly from all other Geomarkets in terms of the substantial racial and ethnic diversity of its purchased profiles. At the lowest PSAT score thresholds (top panel, 1020-1150) the racial/ethnic distribution of prospects within Geomarkets most closely reflects the demographic makeup of the metropolitan area in 2020 (see @fig-chicago-rq1-race). For example, more than 8,000 of the 15,981 total prospects in this SAT range order were concentrated in the Western Suburbs (56% White, 8% Asian, 6% Black, 26% Hispanic, 4% Multiracial) and South and Southwest Suburbs (55% White, 3% Asian, 17% Black, 22% Hispanic, and 3% Multiracial), both of which yield majority White prospect pools comparable to the racial/ethnic makeup of their overall Geomarket populations. The North Shore (78% White, 8% Asian, 2% Black, 8% Hispanic) and the Northwest Suburbs (61% White, 11% Asian, 2% Black, 23% Hispanic) Geomarket pools exhibit the greatest proportions of White prospects. Unlike any other Geomarket, nearly nine of every 10 prospects from the City of Chicago Geomarket were Hispanic (50%), Black (30%), Asian (5%), or Multiracial (2%).

As PSAT score ranges increase, The City of Chicago Geomarket becomes the only Geomarket with substantial shares of Black, and Hispanic prospects. Moving from the top to the bottom panels, all Geomarkets experience growth in the shares of White and Asian prospects, while Black and Hispanic prospect shares decline. For the highest SAT score range, the City of Chicago Geomarket pool is 18% Asian, 7% Black, 21% Hispanic, 5% Multiracial, making it the only Geomarket that is not predominantly White (49%). Prospect pools from all other Geomarkets in this high PSAT order are predominantly White and Asian, with Black and Hispanic shares ranging from 0.4% (North Shore) to 5% (South and Southwest Suburbs) of prospects and 6% (Northwest Suburbs) to 10% (South and Southwest Suburbs) of prospects, respectively. <!--# [LEDE -- BUT IN BOTH THE MEDIUM AND HIGH THRESHOULD, CHICAGO IS THE ONLY GEOMARKET WHERE A SUBSTANTIAL SHARE OF PROSPECTS ARE NOT WHITE OR ASIAN] -->

We also explore contributions across Geomarkets to overall prospect pools by race/ethnicity. @fig-chicago-rq2-midSAT-race shows the contributions of each Geomarket to the pool of prospects from the two orders targeting middle range SAT scores (1160-1300) by race/ethnicity. The top panel presents the full distribution of the 13,643 total prospects (whose race/ethnicity is known) included in these orders by Geomarkets: 11% of all purchased prospect profiles reside in Chain of Lakes, 13% in Northwest Suburbs, 5% in North Shore, 5% in Evanston and Skokie, 17% in City of Chicago, 28% in Western Suburbs, and 20% in South and Southwest Suburbs. These overall distributions are compared to those in the lower panels of @fig-chicago-rq2-midSAT-race, which illustrate the proportional representation of each race/ethnicity group across Geomarkets.

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->
```
@fig-chicago-rq2-midSAT-race shows which Geomarkets send the bulk of students for a given race/ethnicity. The City of Chicago accounts for 17% of all prospects but 48% of Black students. Similarly, South & Southwest Suburbs accounts for 20% of all prospects but 30% of all Black prospects. The upshot is that if the student list purchases had excluded these two Geomarkets, 78% of the nearly 1,000 Black students currently included in these purchases would be excluded. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. For example, the Northwest Suburbs accounts for about 13% of all prospects but only about 3% of Black prospects. The City of Chicago Geomarket also contributed larger shares of the nearly 3,000 Hispanic students (39% versus 17% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

In contrast, these representational patterns by Geomarket are reversed for White and Asian students. @fig-chicago-rq2-midSAT-race shows nearly all Geomarkets contributed to the pool of White prospects at rates proportional to their overall contributions to all included prospects (except City of Chicago). Similar patterns are observed for Asian students, with proportional contributions by Geomarket largely consistent with their overall distribution, except for Northwest Suburbs (19% versus 13% for all prospects) and Evanston and Skokie (10% versus 5% for all prospects) contributing disproportionately larger shares and the South and Southwest Suburbs (8% versus 10% for all prospects) contributing smaller shares relative to their overall contribution to the larger prospect pool.

@@

These contribution patterns by race/ethnicity demonstrate that student list orders targeting prospects in the Chicago metropolitan area using Geomarket filters that exclude the City of Chicago– as observed in some orders collected via our public requests – result in the loss of substantially disproportionate shares of Black and Hispanic students from the resulting prospect pools. Appendix \@fig-chicago-rq2-midSAT-race and @fig-chicago-rq2-highSAT-race show that this pattern holds across both lower-range (1020-1150) and higher-range (1310-1600) SAT score filters. In both cases, the City of Chicago contributes disproportionately larger shares of Black and Hispanic prospects compared to all other Geomarkets. <!--# [REVISE THESE SENTENCES TO HAMMER THE LEDE -- IF WE EXCLUDE CITY OF CHICAGO GEOMAREKT, THE WAY THE ASU ORDER DOES, WE LOSE A DISPROPORTIONATE SHARE OF BLACK AND HISPANIC STUDENTS] -->

We also examine the contributions of each Geomarket to the pool of prospects based on first-generation college student status. Similar to race/ethnicity figures above, @fig-chicago-rq2-midSAT-ses presents the Geomarket distribution for the 13,749 total prospects (whose first-generations status is known) included in middle-range SAT score orders (1160-1300). Among the 1,743 prospects whose parents did not attend college ("no college"), the City of Chicago stands out as the only Geomarket contributing a disproportionately larger share of first-generation college students - 38% in comparison to 17% for all included prospects in this SAT range. In contrast, all other Geomarkets contributed disproportionately smaller shares of first-generation college students whose parents did not attend college relative to all included prospects, with such disparity ranging from a one-percentage-point difference (South and Southwest Suburbs) to a nine-percentage-point difference (Western Suburbs). For prospects whose parents attended some college but did not complete their degrees ("some college"), @fig-chicago-rq2-midSAT-ses indicates both City of Chicago and South and Southwest Suburbs Geomarkets contribute a disproportionately larger share of first-generation students. These representational patterns are reversed among prospects that are not first-generation college students. The City of Chicago Geomarket contributed a substantially smaller share of not-first generation prospects (12%) relative to all prospects (17%) in this SAT range, whereas all other Geomarkets contributed comparable shares.

@@

Finally, we examine Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Chicago metropolitan area. @fig-chicago-rq2-midSAT-combo presents the Geomarket distribution of the 12,981 total prospects (whose first-generation status and race/ethnicity is known) included in middle-range SAT (1160-1300) orders. @fig-chicago-rq2-midSAT-combo shows the City of Chicago Geomarket consistently contributes a disproportionately larger share of first-generation college students whose parents did not attend college across all racial/ethnic categories. For instance, among all Asian prospects (n=1,341) whose contact information was purchased in this SAT score range, approximately 12% were first-generation college students whose parents did not attend college, 16% were first-generation college students whose parents attended some college but did not complete their degree, and 72% had parents with a BA. By contrast, for 237 Asian students from the City of Chicago Geomarket, 35% had parents who did not attend colllege and only 43% had parents with a BA.

Similar Geomarket contribution patterns are evident among Hispanic prospects. The City of Chicago Geomarket stands out as the only Geomarket contributing substantially larger shares of Hispanic first-generation college students across those whose parents did not attend college (46% versus 38% overall) and those whose parents attended some college but did not complete their degrees (29% versus 26% overall). By contrast, Hispanic students from the more affluent North Shore Geomarket and Evanston & Skokie Geomarket were more likely to have parents with a BA.

Although Black prospects who are not first-generation college students make up a relatively smaller portion of the overall pool, @fig-chicago-rq2-midSAT-combo suggests that such prospects are primarily contributed by the Evanston and Skokie, City of Chicago, and Western Surburbs Geomarkets. Similarly, Geomarkets such as Chain of Lakes, North Shore, Evanston and Skokie, and Western Surburbs contribute a disproportionately larger share of White prospects who are not first-generation college students, whereas City of Chicago and South and Southwest Suburbs contribute larger shares of White first-generation college students.

@@

```{r fig-chicago-rq2-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Purchased Student Profiles by Geomarket, Chicago Area", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_chicago_race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq2-midSAT-race, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by Racial/ethnic group, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq2-midSAT-ses, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_ses.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq2-midSAT-combo, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_combo.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

>>>>

@fig-chicago-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Chicago metropolitan area, notably revealing that the City of Chicago stands out distinctly from all other Geomarkets in terms of the substantial racial and ethnic diversity of its purchased profiles. At the lowest PSAT score thresholds (top panel, 1020-1150) the racial/ethnic distribution of prospects within Geomarkets most closely reflects the demographic makeup of the metropolitan area in 2020 (see @fig-chicago-rq1-race). For example, more than 7,000 of the 15,964 total prospects in this SAT range order were concentrated in the Western Suburbs (59% White, 9% Asian, 6% Black, 23% Hispanic, 4% Multiracial) and South and Southwest Suburbs (54% White, 2% Asian, 19% Black, 21% Hispanic, and 3% Multiracial), both of which yield majority White prospect pools comparable to the racial/ethnic makeup of their overall Geomarket populations. The North Shore (70% White, 12% Asian, 4% Black, 11% Hispanic) and the Northwest Suburbs (59% White, 11% Asian, 2% Black, 24% Hispanic) Geomarket pools exhibit the greatest proportions of White prospects. Unlike any other Geomarket, nearly eight of every 10 prospects from the City of Chicago Geomarket were Hispanic (47%), Black (24%), Asian (5%), or Multiracial (3%).

As PSAT score ranges increase, The City of Chicago Geomarket becomes the only Geomarket with substantial shares of Black and Hispanic prospects. Moving from the top to the bottom panels, all Geomarkets experience growth in the shares of White and Asian prospects, while Black and Hispanic prospect shares decline. For the highest SAT score range, the City of Chicago is the only Geomarket where White prospects comprise a slim majority (55%), with the remaining prospect pool made up of 16% Asian, 6% Black, 18% Hispanic, 5% Multiracial students. Prospect pools from all other Geomarkets in this high PSAT order are predominantly White and Asian, with Black and Hispanic shares ranging from 0.3% (North Shore) to 4% (South and Southwest Suburbs) of prospects and 5% (North Shore) to 10% (South and Southwest Suburbs) of prospects, respectively. <!--# [LEDE -- BUT IN BOTH THE MEDIUM AND HIGH THRESHOULD, CHICAGO IS THE ONLY GEOMARKET WHERE A SUBSTANTIAL SHARE OF PROSPECTS ARE NOT WHITE OR ASIAN] -->

We also explore contributions across Geomarkets to overall prospect pools by race/ethnicity. @fig-chicago-rq2-midSAT-race shows the contributions of each Geomarket to the pool of prospects from the two orders targeting middle range SAT scores (1160-1300) by race/ethnicity. The top panel presents the full distribution of the 13,729 total prospects (whose race/ethnicity is known) included in these orders by Geomarkets: 10% of all purchased prospect profiles reside in Chain of Lakes, 11% in Northwest Suburbs, 6% in North Shore, 4% in Evanston and Skokie, 23% in City of Chicago, 29% in Western Suburbs, and 17% in South and Southwest Suburbs. These overall distributions are compared to those in the lower panels of @fig-chicago-rq2-midSAT-race, which illustrate the proportional representation of each race/ethnicity group across Geomarkets.

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->
```
@fig-chicago-rq2-midSAT-race shows which Geomarkets send the bulk of students for a given race/ethnicity. The City of Chicago accounts for 23% of all prospects but 51% of Black students. Similarly, South & Southwest Suburbs accounts for 17% of all prospects but 26% of all Black prospects. The upshot is that if the student list purchases had excluded these two Geomarkets, 77% of the nearly 1,000 Black students currently included in these purchases would be excluded. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. For example, the Northwest Suburbs accounts for about 11% of all prospects but only about 3% of Black prospects. The City of Chicago Geomarket also contributed larger shares of the more than 2,500 Hispanic students (45% versus 23% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

In contrast, these representational patterns by Geomarket are reversed for White and Asian students. @fig-chicago-rq2-midSAT-race shows nearly all Geomarkets contributed to the pool of White prospects at rates proportional to their overall contributions to all included prospects (except City of Chicago). Similar patterns are observed for Asian students, with proportional contributions by Geomarket largely consistent with their overall distribution, except for Northwest Suburbs (17% versus 11% for all prospects) and Evanston and Skokie (9% versus 4% for all prospects) contributing disproportionately larger shares and the South and Southwest Suburbs (6% versus 17% for all prospects) contributing smaller shares relative to their overall contribution to the larger prospect pool.

@@

These contribution patterns by race/ethnicity demonstrate that student list orders targeting prospects in the Chicago metropolitan area using Geomarket filters that exclude the City of Chicago– as observed in some orders collected via our public requests – result in the loss of substantially disproportionate shares of Black and Hispanic students from the resulting prospect pools. The [Online Appendix](https://ozanj.github.io/cb_geo/#/results) shows that this pattern holds across both lower-range (1020-1150) and higher-range (1310-1600) SAT score filters. In both cases, the City of Chicago contributes disproportionately larger shares of Black and Hispanic prospects compared to all other Geomarkets. <!--# [REVISE THESE SENTENCES TO HAMMER THE LEDE -- IF WE EXCLUDE CITY OF CHICAGO GEOMAREKT, THE WAY THE ASU ORDER DOES, WE LOSE A DISPROPORTIONATE SHARE OF BLACK AND HISPANIC STUDENTS] -->

We also examine the contributions of each Geomarket to the pool of prospects based on first-generation college student status. Similar to race/ethnicity figures above, @fig-chicago-rq2-midSAT-ses presents the Geomarket distribution for the 13,848 total prospects (whose first-generations status is known) included in middle-range SAT score orders (1160-1300). Among the 1,750 prospects whose parents did not attend college ("no college"), the City of Chicago stands out as the only Geomarket contributing a disproportionately larger share of first-generation college students - 43% in comparison to 23% for all included prospects in this SAT range. In contrast, all other Geomarkets contributed disproportionately smaller shares of first-generation college students whose parents did not attend college relative to all included prospects, with such disparity ranging from a one-percentage-point difference (Northwest Suburbs, South and Southwest Suburbs) to a ten-percentage-point difference (Western Suburbs). For prospects whose parents attended some college but did not complete their degrees ("some college"), @fig-chicago-rq2-midSAT-ses indicates both City of Chicago and South and Southwest Suburbs Geomarkets contribute a disproportionately larger share of first-generation students. These representational patterns are reversed among prospects that are not first-generation college students. The City of Chicago Geomarket contributed a substantially smaller share of not-first generation prospects (18%) relative to all prospects (23%) in this SAT range, whereas all other Geomarkets contributed comparable shares.

@@

Finally, we examine Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Chicago metropolitan area. @fig-chicago-rq2-midSAT-combo presents the Geomarket distribution of the 13,059 total prospects (whose first-generation status and race/ethnicity is known) included in middle-range SAT (1160-1300) orders. @fig-chicago-rq2-midSAT-combo shows the City of Chicago Geomarket consistently contributes a disproportionately larger share of first-generation college students whose parents did not attend college across all racial/ethnic categories. For instance, among all Asian prospects (n=1,370) whose contact information was purchased in this SAT score range, approximately 12% were first-generation college students whose parents did not attend college, 16% were first-generation college students whose parents attended some college but did not complete their degree, and 73% had parents with a BA. By contrast, for 292 Asian students from the City of Chicago Geomarket, 30% had parents who did not attend college and less than 50% had parents with a BA.

For Hispanic prospects, the City of Chicago stands out as the Geomarket contributing the largest overall number and share of Hispanic first-generation college students across those whose parents did not attend college (44% versus 38% overall) and those whose parents attended some college but did not complete their degrees (29% versus 26% overall). Other Geomarkets such as Northwest Suburbs and Chain of lakes also contributed substantially larger shares. By contrast, Hispanic students from the more affluent North Shore Geomarket and Evanston & Skokie Geomarket were more likely to have parents with a BA.

Although Black prospects who are first-generation college students make up a relatively smaller portion of the overall pool, @fig-chicago-rq2-midSAT-combo suggests that such prospects are primarily contributed by the Evanston and Skokie, City of Chicago, and South and Southwest Surburbs Geomarkets. Similarly, Geomarkets such as Chain of Lakes, North Shore, Evanston and Skokie, and Western Surburbs contribute a disproportionately larger share of White prospects who are not first-generation college students, whereas City of Chicago and South and Southwest Suburbs contribute larger shares of White first-generation college students.

@@

```{r fig-chicago-rq2-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Purchased Student Profiles by Geomarket, Chicago Area", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq2/rq2_chicago_race_row_plot.png')
#knitr::include_graphics('./../assets/images/rq2_chicago_race.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq2-midSAT-race, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by Racial/ethnic group, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq2/rq2_chicago_race_col_plot_392834_488035.png')

#knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_race.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq2-midSAT-ses, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq2/rq2_chicago_firstgen_col_plot_392834_488035.png')
#knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_ses.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq2-midSAT-combo, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
#
knitr::include_graphics('./../results/graphs/rq2b/rq2b_chicago_order_392834_488035_race_by_firstgen.png')
#knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_combo.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
